---
title: "Accessing EMF public spatial products with R"
author:
  - vgranda
  - emf
format:
  hugo-md:
    mermaid-format: svg
    keep-yaml: false
    preserve-yaml: false
editor_options: 
  chunk_output_type: console
---

## Introduction

Along with the main activities we at the EMF work on, including the development and
maintenance of the web apps in the [LFC](https://laboratoriforestal.creaf.cat/),
we generate several data products. This products are made publicly available for
anyone to use them.

### Datasets available

  - [Daily continuous meteorological data (500 m²) for peninsular Spain and 
  Balearic Islands](https://emf.creaf.cat/data/meteoland_spain/).
  This data is interpolated from meteorological stations using the
  [`meteoland` R package](https://emf.creaf.cat/software/meteoland/).  
  Data format is `gpkg`, containing points in a 500x500 grid in the ETRS89 / UTM
  zone 30N coordinate reference system (points represent the cell center).  
  Each file contains ~2 million rows and an avergage file size of 630MB.

  - [Daily continuous modelled forest water balance for peninsular Spain and
  Balearic Islands](https://emf.creaf.cat/data/medfateland_spain/).
  Data has been generated by the
  [MEDFATE model](https://emf.creaf.cat/models/medfate_model/)
  from soil, climate and forest inputs.  
  Data format is `gpkg`, containing points in a 500x500 grid in the ETRS89 / UTM
  zone 30N coordinate reference system (points represent the cell center).  
  Each file contains ~760000 rows and an average file size of 175MB

### Accessing the data

The simplest way of using the data is downloading the desired files from the
public reposiory (https://data-emf.creaf.cat/public/gpkg/). But because the size
of the files, this can be time, memory and disk storage consuming.  
Given the formats offered, in this case we can make use of the capabilities
of the [GDAL system library](https://gdal.org/en/latest/index.html) and its
implementation in R by the `sf` package to avoid this costs.

## Simple example

If we only need to access one day, the simplest way is to use `sf` to read the
file directly from the server.

```{r}
#| cache: true
# libraries needed
library(sf)
library(dplyr)
library(mapSpain)
library(glue)

# retrieving 25th April 2025
data_20250425 <- st_read(
  "https://data-emf.creaf.cat/public/gpkg/daily_interpolated_meteo/20250425.gpkg"
)

# explore result
data_20250425
```

But this is not so different from downloading the data and reading it 
locally, as the default behaviour of `st_read` is to download the file to a
temporal folder and read it from there.  
It's also not so efficient in terms of
time, as we need to wait to download the file and read all the data, which also
means storing all data in memory.

## Filtering by polygon

If we don't need all the data extent (peninsular Spain and Balearic Islands), but
an smaller area instead, we can take advantage of the `wkt_filter` argument in
`st_read`.  
Given a polygon (in *Well Known Text* format), `st_read` uses GDAL to
access and retrieve only the part of the files that intersects with that
polygon. This is useful because reduces transfer size (only a part of the file
is read and transfered), time and memory used.  
For this, before reading the remote file, we need to get the polygon of the area
we want. In the following example we'll use the
[`mapSpain`](https://ropenspain.github.io/mapSpain/) R package to obtain the
polygon corresponding to Posada de Valdeón (Léon) municipality and use it to
retrieve meteorological data for that polygon from the same file as before:

```{r}
#| cache: true
posada_wkt <- mapSpain::esp_get_munic(munic = "Posada de Valdeón") |>
  # ensure the poly is in the same CRS that the data
  sf::st_transform(crs = 25830) |>
  # get the polygon geometry
  dplyr::pull("geometry") |>
  # transform to text (Well Known Text, wkt)
  sf::st_as_text()

# retrieving 25th April 2025 only for Posada
data_posada_20250425 <- st_read(
  "https://data-emf.creaf.cat/public/gpkg/daily_interpolated_meteo/20250425.gpkg",
  wkt_filter = posada_wkt
)

# explore result
data_posada_20250425
```

:::{.callout-important}
Be sure that the polygon is in the **same** Coordinate Reference System (CRS)
as the data (EPSG:25380). `wkt_filter` don't check this and assumes the filter
polygon is in the same CRS, so most probably it will return no data when a
different CRS is used.
:::

## Filtering by other variables (with queries)

Another option we have to reduce the data read into memory is retrieving only
the data we are really interested on. Besides the spatial filter we saw in the
previous section, GDAL also exposes an SQL dialect that can be
used to query only the parts of the data we need. This means we can only read
the columns and rows that comply with the supplied query.  
We will use the `query` argument in `st_read` to make use of this GDAL feature
and read only temperatures data in Posada de Valdeón for points above
1500m asl:

```{r}
#| cache: true
# creating the query text
query_text <- glue::glue(
  "SELECT
    geom, elevation, MeanTemperature, MinTemperature, MaxTemperature
  FROM '20250425'
  WHERE elevation >= 1500;"
)

# retrieving 25th April 2025 temperatures only for Posada >= 1500m
data_posada_1500m_20250425 <- st_read(
  "https://data-emf.creaf.cat/public/gpkg/daily_interpolated_meteo/20250425.gpkg",
  wkt_filter = posada_wkt,
  query = query_text
)

# explore result
data_posada_1500m_20250425
```

Let's see the query with more detail:

  - `SELECT geom, elevation, MeanTemperature, MinTemperature, MaxTemperature`
  selects the specified data columns. If we want to retrieve all columns, we can
  use `SELECT *`, were `*` means all columns.

  - `FROM '20250425'` the table, in this case, the name of the file without the
  extension.

  - `WHERE elevation >= 1500` the filters we apply. If we apply more than one
  filter we need to use logical connectors, like
  `WHERE elevation >= 1500 AND MinTemperature < 3`

If both arguments, `query` and `wkt_filter` are used, the geometry column must
be always selected (`SELECT geom` in this example) or the wkt_filter will not
work.

:::{.callout-tip}
For more info about the GDAL SQL dialect, visit
https://gdal.org/en/stable/user/ogr_sql_dialect.html
:::

## Temporal series (loop)

All the previous examples are using only one file (day). If we want to read
more than one, then the simplest way is to make a loop for all the dates we
want to work with

```{r}
#| cache: true
# create the dates to read, but in the format of the files in the server
dates_to_read <-
  seq(as.Date("2025-04-25"), as.Date("2025-04-27"), by = "day") |>
  as.character() |>
  stringr::str_remove_all("-")

# loop by dates
data_posada_days <- purrr::map(
  dates_to_read,
  \(day) {
    # query for the day
    query_text <- glue::glue(
      "SELECT
        geom, dates, elevation, MeanTemperature, MinTemperature, MaxTemperature
      FROM '{day}'
      WHERE elevation >= 1500;"
    )
    # read the posada data of that day
    st_read(
      glue::glue("https://data-emf.creaf.cat/public/gpkg/daily_interpolated_meteo/{day}.gpkg"),
      wkt_filter = posada_wkt,
      query = query_text
    )
  }
) |>
  # bind the results in a dataframe
  purrr::list_rbind()

# explore result
head(data_posada_days)
```

